---
title: "HW2 NRI Risk Ranking"
author: "Stephan Kadonoff"
format: html
editor: visual
date: 1/27/2025
execute: 
  message: false
  warning: false
---

## National Risk Index Analysis

```{r}
# initial data input and wrangling
library(tidyverse)
library(here)
library(janitor)

nri_data = read_csv(here("data", "National_Risk_Index_Counties_807384124455672111.csv"))

# clean the column names and filter out for only the 50 states
nri_data_clean <- nri_data %>%
  clean_names()

```

```{r}
#| fig-height: 14
#| fig-width: 9
#| fig-alt: "Boxplots of National Risk Index Scores for all 50 states. California highlighted for comparison."
# create a boxplot to visualize the national risk index scores per state

nri_data_clean %>%
  filter(state_name %in% state.name) %>%
  mutate(is_ca = state_name == "California") %>% # isolate California for visualization
  ggplot(aes(x = reorder(state_name, national_risk_index_score_composite, median),
             y = national_risk_index_score_composite,
             fill = is_ca)) + 
  geom_boxplot() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "dodgerblue", "FALSE" = "grey")) +
  labs(title = "National Risk Index Scores by State",
       subtitle = "California displays highest median risk index score",
       caption = "Source: FEMA National Risk Index",
       x = NULL,
       y = "Risk Index Score") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))


  
```

1.  What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?

My variables of interest are numeric and categorical. We are essentially finding the range risk index scores for each state's respective counties, summing it, and then plotting them based on their average. Therefore, the state names are categorical, but the actual index scores are numeric.

2.  How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?

Since I knew that we had to visualize all 50 states, and have them ordered in such a way that each is equally visualized in comparison with one another save for California, I decided a blox plot would be a good fit. I decided on a boxplot because the data rows are at the county level. Since there are so many county rows per each state, I figured it would be a good way to show the their overall spread in comparison with one another. I think a heat-map, a ribbon plot, or some other plot that can show many different ranges on the same plot would be ideal for this.

3.  Summarize your main finding in no more than two sentences.

California has the highest risk index score on average. When you compare it to the other large states, such as Texas, New York, and even Florida, California has a much higher risk rate then one would expect.

4.  What modifications did you make to this visualization to make it more easily readable?

I colored California's specific boxplot. I also ordered the plot such that the highest index scores would be at the top.

5.  Is there anything you wanted to implement, but didn’t know how? If so, please describe.

I think the part of plotting is just getting more exposure to a wide breath of plots, and how best to know which plot type fits your data the best
